|

### 2020-04-28 documents/ovando-et-al-pop-effects-of-mpas-si.Rmd
a bit of cleanup

---
title: "Supporting Information"
subtitle: "Assessing the Population-level Conservation Effects of Marine Protected Areas"
author: 
  - Daniel Ovando
  - Jennifer E. Caselle
  - Christopher Costello
  - Olivier Deschenes
  - Steven D. Gaines
  - Ray Hilborn
  - Owen Liu
output: 
  bookdown::pdf_document2:
    keep_tex: true
  pdf_document: default
bibliography: My Library.bib
header-includes:
    - \usepackage{longtable}
    - \usepackage{booktabs}
    - \usepackage{eso-pic}
    - \usepackage{graphicx}
    - \usepackage[left]{lineno}
    - \usepackage{xcolor}
params:
  run_name: ["v6.0"]
  sim_years: [50]
  burn_years: [20]
  num_patches: [25]
always_allow_html: true
---

@@

\newpage
```{r cartoon, fig.cap= "Cartoon illustration of the hierarchichal difference-in-difference estimator", fig.width=3}
knitr::include_graphics("figs/estimation-cartoon.png")
```

>>>>

---
title: "Supporting Information - Appendix"
subtitle: "Assessing the Population-level Conservation Effects of Marine Protected Areas"
author: 
  - Daniel Ovando
  - Jennifer E. Caselle
  - Christopher Costello
  - Olivier Deschenes
  - Steven D. Gaines
  - Ray Hilborn
  - Owen Liu
output: 
  bookdown::pdf_document2:
    keep_tex: true
  pdf_document: default
bibliography: My Library.bib
header-includes:
    - \usepackage{longtable}
    - \usepackage{booktabs}
    - \usepackage{eso-pic}
    - \usepackage{graphicx}
    - \usepackage[left]{lineno}
    - \usepackage{xcolor}
params:
  run_name: ["v6.0"]
  sim_years: [50]
  burn_years: [20]
  num_patches: [25]
always_allow_html: true
---

@@

\newpage
```{r cartoon, fig.cap= "Cartoon illustration of the hierarchichal difference-in-difference estimator", fig.width=8}
knitr::include_graphics("figs/estimation-cartoon.png")
```

### 2020-04-28 documents/ovando-et-al-pop-effects-of-mpas.Rmd
a bit of cleanup

---
title: |
  | Assessing the Population-level Conservation Effects of
  | Marine Protected Areas
author: 
  - Daniel Ovando
  - Jennifer E. Caselle
  - Christopher Costello
  - Olivier Deschenes
  - Steven D. Gaines
  - Ray Hilborn
  - Owen Liu
abstract: |
   Marine Protected Areas (MPAs) cover 3-7% of the world’s ocean, up from less than 1% in the year 2000, and international commitments call for 10%, 30%, and even 50% coverage. The premise underlying MPA expansion is that they conserve biodiversity, habitats, and fished populations. While numerous studies show that MPAs produce conservation benefits *inside* their borders, many MPAs are also justified on the grounds that they confer conservation benefits to the broader population *beyond* their borders. This paper examines the conditions under which MPAs can provide population-level conservation benefits inside and outside their borders, and shows that even in cases where the conservation benefits are large, they are inherently difficult to detect empirically. A network of MPAs was put in place in The Channel Islands National Marine Sanctuary in 2003, with a goal of providing regional conservation and fishery benefits. Evidence indicates that the Channel Island MPAs have increased biomass densities inside the MPAs, but we are unable to find a clear effect of these same MPAs at the population level. MPA effect sizes less than 30% are likely to be difficult to detect (even when they are present), and the size of many MPA networks suggests that effect sizes may often be smaller than 30%. Our results provide a novel assessment of the population-level effects of a large and iconic Marine Protected Area network, and provide guidance for communities charged with monitoring and adapting MPAs.
bibliography: My Library.bib
csl: pnas.csl
output:
  bookdown::pdf_document2: default
  bookdown::word_document2: default
  bookdown::html_document2: default
params:
  run_name: ["v6.0"]
  sim_years: [50]
  burn_years: [20]
  num_patches: [25]
linkcolor: blue
toc: false
header-includes:
  - \usepackage{setspace}\doublespacing
  - \usepackage{lineno}\linenumbers
---

```{r, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE, dpi = 600,
                      cache = FALSE, fig.width = 3.42, fig.asp = 1.5, dev = "cairo_pdf")

@@

```{r, include=FALSE}
library(sf)
library(ggmap)
library(viridis)
library(hrbrthemes)
library(scales)
library(patchwork)
library(rpart)
library(extrafont)
library(caret)
library(ggsci)
library(rEDM)
library(tidyverse)
library(ggtext)
extrafont::loadfonts()
functions <- list.files(here::here("functions"))

@@

@@

# Body

@@

Given this existing body of theory, what should we expect *a priori* the effect size of the Channel Islands MPAs to be? While we know many parameters will affect the expected effects of the MPAs, theory suggests that life history, adult and larval dispersal relative to MPA size, and fishing pressure will be key drivers. The targeted species in our database span a range of life histories, but are largely made up of fishes in the perch and rockfish complexes, with a mean Von Bertalanffy growth coefficient of `r round(mean(used_data$k),2)`, and a median age at maturity of `r round(median(used_data$tm),0)` years [values drawn from @thorson2017c].

@@

To illustrate this problem, we simulated response ratio trajectories for species and MPA coverage representative of the Channel Islands (see methods and SI for details of model structure). These simulations cover the life-histories and MPA sizes seen in the Channel Islands, but vary in key unknowns such as the actual degree of fishing pressure, the timing of density dependence, and the fishing fleets response to the MPAs. For each year of protection, we paired our simulated response ratios to the estimated posterior probability distributions of the  empirical Channel Island response ratios. For example, if the mean estimated response ratio in the year 2006 (year three after implementation) is one, we found all simulations that had simulated response ratios near one after three years of protection, and then pulled out the "true" simulated population-level MPA effect from each of those simulations. This provides us with a distribution of simulated population-level MPA effects that could plausibly generate the types of response ratios actually observed in the Channel Islands.

@@

ggsave(
  temp_plot,
  filename = here::here("documents","figs","response_ratio_plot.png"),
  width = width_1col,
  height = width_1col * asp_1col,
  dpi = dpi
)

@@

ggsave(
  mpa_effect_plot,
  filename = here::here("documents","figs","mpa_effect_plot.png"),
  width = width_2col,
  height = width_2col * asp_2col,
  dpi = dpi
)

@@

While we do not have access to fine scale fishing data from the Channel Islands alone, reported catches for the species of interest in the Santa Barbara region exhibit a mix of stable, downward, and upward trajectories (see Fig.S1) which indicates that a negative MPA effect caused by a constant-catch fishing strategy is unlikely. However, some of the most common targeted species in the Channel Islands, such as blue rockfish and copper rockfish, have seen dramatic increases in reported catches in the Santa Barbara region since the year 2010 (although other such as California sheephead have seen declines relative to pre-MPA catch level, Fig.S1). To the extent that these increases in catches are due to external market forces, the dissapearance of clear positive population-level MPA effects estimated by our model could be due to exogenous increases in fishing pressure overwhelming any population-level increases caused by the MPA network. We lack a control though that would allow us to estimate the MPA effects relative to a counter-factual in which the same market forces increase catches but no MPAs are in place.

@@

This hypothesis is supported by the fact that we estimate a similar decline in the estimated MPA effect both inside and outside of the MPAs (Fig.S18) (as well as observe the decline in the raw data, Fig.\@ref(fig:raw-density)). If the cause of recent declines was due to increases in fishing pressure we would expect to see substantial declines only in the fished areas. While we see evidence for the validity of the parallel trends assumption in the early years of our data (Fig.\@ref(fig:mpa-effect), Fig.S26), it is likely that the non-targeted species reacted in a different way to the marine heatwave than the targeted species did, causing a violation of the parallel trends assumption in the post 2014 data.

ggsave(
temp_plot,
filename = here::here("documents","figs","raw_biomass_density_plot.png"),
  width = width_1col,
  height = width_1col * asp_1col,
    dpi = dpi
)

@@

```{r}
tmdata <- life_history_data %>% 
  filter(targeted == 1, classcode %in% top_species) %>% 
  select(tm, everything())

median_tm = round(median(tmdata$tm))
```

Our simulations and difference-in-difference model examine percentage changes in biomass densities (as opposed to changes in total population size). An increase in biomass densities from 0.02kg/m^2^ to 0.04kg/m^2^ would be measured as a 100% increase. While a large percentage effect, this is a small change in biomass densities relative to the variance in the observation process itself. While an MPA providing 25% protection might be expected to produce a 30% increase in biomass densities, expressed as a percentage of unfished biomass, this gain is on the order of 15% (Fig.S45). In addition, the median estimated age at sexual maturity for the targeted species included in this study is `r median_tm` years, meaning that the span of this analysis represents less than three generations of MPA protection for half of the measured species. Ongoing monitoring may yet reveal clearer effects, especially if the time series extends beyond the periods of environmental shocks. Analysis of more rapidly growing and maturing species, e.g. spiny lobster, may also reveal clearer signals.

@@

temp_plot
ggsave(
  temp_plot,
  filename = here::here("documents","figs","validation_plot.png"),
  width = width_1col,
  height = width_1col * asp_1col,
    dpi = dpi

When should we expect to see MPA effects big enough to stand a reasonable chance of detection? We simulated `r n_distinct(outcomes$experiment)` MPA scenarios across a wide range of life histories, network designs, and fishing dynamics (see SI). Suppose that we are willing to tolerate a MAPE of 25%. Our analysis suggests that we would need a population-level MPA effect size of at least 30% to achieve this. Two of the most critical drivers of MPA effect size are the size of the MPA network and the degree of fishing pressure. Looking across these two variables, if the MPA network covers 25% or more of a species range and pre-MPA depletion is greater than 60% we might expect an effect size with our target MAPE. Although recently some extremely large MPAs have been enacted that may indeed reach into the higher levels of MPA coverage, most MPA networks for near-shore commercial finfish are likely to cover areas more in line with the Channel Islands (20%) or smaller, and as such have population-level effect sizes that may be difficult to detect unless they were experiencing extreme overfishing prior to MPA establishment (Fig.\@ref(fig:sim-effects)-A).

@@

To put these results into context, the FAO estimates that 7% of the worlds fisheries with status estimates fall into the relatively unexploited category (roughly depletion less than 50%), 60% fall into the fully fished category (roughly depletion 50%-70%), and 33% fall into the heavily fished category (roughly depletion greater than 70%) [@fao2018, assuming a biomass at maximum sustainable yield to unfished biomass ratio of 40%, per @thorson2012], although studies that include a broader range of fisheries estimate that 50% or more of stocks fall into the heavily fished category [@costello2012;@rosenberg2018]. The population-level effects of MPA networks covering 25% or less of a species range may be difficult to detect in many places with already well-managed fisheries, while for that MPA network size we might expect clearer effects in less-managed locations (assuming compliance with MPA regulations). Within these broad guidelines a wide range of outcomes are possible based on local fleet and fish dynamics: as a starting place users can use the bio-economic MPA simulation model developed for this paper to explore potential outcomes for specific MPAs using an interactive web application available [here](https://danovando.shinyapps.io/simmpa/).

expected_mpa_effect_plot
ggsave(
  expected_mpa_effect_plot,
  filename = here::here("documents","figs","expected_mpa_effect.png"),
  width = width_2col,
  height = width_2col * asp_2col,
    dpi = dpi

@@

Non-equilibrium analyses also help set expectations for effect sizes over time [@nickols2019,@kaplan2019,Fig.\ref{did-plot}]. Educating communities about the challenges of estimating the effects of MPAs can help set expectations, so that a lack of a clear effect is not necessarily viewed as a failure of the program, or large positive result based solely on response ratios as a clear sign of success and subsequent relaxation of other fishery management strategies. Rather, results and subsequent management actions (such as adaptation of MPA networks) must be considered in the context of reasonable expectations given the size, age, and degree of enforcement of the MPAs in question, together with the ecological and economic dynamics of a given system. While this paper has focused on the conservation outcomes of MPAs, future work must also address the challenge of predicting and estimating the fishery impacts of protected areas [in the manner of @smith2006a].

@@

We present here critical characteristics of our simulation model and regression approach. Further details can be found in the Supplementary Information. All analysis were conducted in R [@rcoreteam2019]. Our main difference-in-difference model was fit using Stan [@carpenter2017] using the `rstanarm` package [@goodrich2020]. All materials needed to reproduce results can be found [here](https://github.com/DanOvando/regional-effects-of-mpas).

@@

Our bio-economic model simulates the effect of MPAs on a spatially explicit age-structured representation of a single species. Readers can explore the functionality of the model using an online tool available [here](https://danovando.shinyapps.io/simmpa/). Full details of the simulation model are available in the Supplementary Information.

@@

temp_plot
ggsave(
 temp_plot,
  filename = here::here("documents","figs","sample_location_plot.png"),
  width = width_1col,
  height = width_1col * asp_1col,
    dpi = dpi

@@

The key assumptions of the difference-in-difference model are that a) within the time-frame of the model there are no significant interaction effects between the targeted and non-targeted species (which in fact we do not detect, see SI), and b) that in the absence of the MPAs both the targeted and non-targeted groups of species would have exhibited similar trends in densities. It is important to note that the model is agnostic as to the root causes of the parallel trends, so long as they hold.

@@

where $d_i$ is the biomass density at observation *i*, *T* indicates whether the observation *i* is for a targeted ($T = 1$) or non-targeted ($T = 0$) species, and *MPA* marks whether observation *i* is in a pre MPA ($MPA = 0$) or post MPA ($MPA = 1$) state. **B^c^** is a vector of coefficients for additional control variables in matrix *X* such as water visibility and observer experience. **B^s^** is a vector of hierarchical coefficients for each sampling location **S** , clustered by island.  Under the assumptions of this model, $\beta_3$ is the causal effect of the treatment (*MPA*) on the treated (targeted species). $\beta_0$ is the mean total biomass of non-targeted species pre-MPA, and $\beta_1T$ is the mean total biomass of targeted species pre-MPA. $\beta_2MPA$ is the effect of the post-MPA period on the non-targeted species. The shape and scale parameters of the Gamma distribution are estimated as well. See SI for further details of the estimation model. `r percent(mean(used_data$eventual_mpa))` of the samples in the raw data come from within MPAs, while the MPAs themselves cover 20% of the surface waters of the Channel Islands. As such, we weight each observation such that within MPA data are assigned a total weight of 0.2 and outside-MPA data a total weight of 0.8. See SI for sensitivity analyses of various model structures and synthetic controls results.

We briefly assess two of the most critical assumptions of this model here: that the treated and non-treated groups have parallel trends, and that the effect of the treatment on the treated does not tangentially affect the untreated. While the parallel trends assumption cannot be formally proven, we can examine its validity using the data from the years before the MPAs were put in place in 2003. We do not detect any significant differences in the trends of the biomass densities of the targeted and non-targeted species in the years before the MPAs (Fig.S26). However, we believe that the parallel trends assumption is likely violated in the post 2014 data.

With regards to the second assumption, all of the species in this empirical analysis exist within an ecosystem, and as such affect each other through mechanisms such as predation, competition, and habitat modification. We find it unlikely that these effects have had enough time to manifest in a meaningful way in the 14 years of post-MPA data used in our analysis [@babcock2010; @pershing2015a]. We used convergent cross mapping (CCM), in the manner of @clark2015, to test for the possibility of the trophic cascades biasing our results. Our results found no significant cross-mappings between targeted and non-targeted species, indicating that while clearly there are interactions between these groups on some level, the effects within the timespan of the data are not pronounced enough to be of concern to our results (Fig.S42). However, the longer MPAs are in place, the greater the possibility that substantial species interactions that can affect use of non-targeted species as a control may arise.

>>>>

---
title: |
  | Assessing the Population-level Conservation Effects of
  | Marine Protected Areas
author: 
  - Daniel Ovando
  - Jennifer E. Caselle
  - Christopher Costello
  - Olivier Deschenes
  - Steven D. Gaines
  - Ray Hilborn
  - Owen Liu
abstract: |
   Marine Protected Areas (MPAs) cover 3-7% of the world’s ocean, up from less than 1% in the year 2000, and international commitments call for 10%, 30%, and even 50% coverage. The premise underlying MPA expansion is that they conserve biodiversity, habitats, and fished populations. While numerous studies show that MPAs produce conservation benefits *inside* their borders, many MPAs are also justified on the grounds that they confer conservation benefits to the broader population *beyond* their borders. This paper examines the conditions under which MPAs can provide population-level conservation benefits inside and outside their borders, and shows that even in cases where the conservation benefits are large, they are inherently difficult to detect empirically. A network of MPAs was put in place in The Channel Islands National Marine Sanctuary in 2003, with a goal of providing regional conservation and fishery benefits. Evidence indicates that the Channel Island MPAs have increased biomass densities inside the MPAs, but we are unable to find a clear effect of these same MPAs at the population level. MPA effect sizes less than 30% are likely to be difficult to detect (even when they are present), and the size of many MPA networks suggests that effect sizes may often be smaller than 30%. Our results provide a novel assessment of the population-level effects of a large and iconic Marine Protected Area network, and provide guidance for communities charged with monitoring and adapting MPAs.
bibliography: My Library.bib
csl: pnas.csl
output:
  bookdown::word_document2:
    reference_docx: template.docx
  bookdown::pdf_document2: default
  bookdown::html_document2: default
params:
  run_name: ["v6.0"]
  sim_years: [50]
  burn_years: [20]
  num_patches: [25]
linkcolor: blue
toc: false
header-includes:
  - \usepackage{setspace}\doublespacing
  - \usepackage{lineno}\linenumbers
---

```{r, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE, dpi = 600,
                      cache = FALSE, fig.width = 3.42, fig.asp = 1.5, dev = "png")

@@

```{r, include=FALSE}
library(sf)
library(ggmap)
library(viridis)
library(hrbrthemes)
library(scales)
library(patchwork)
library(rpart)
library(extrafont)
library(caret)
library(ggsci)
library(rEDM)
library(tidyverse)
library(ggtext)
library(here)
extrafont::loadfonts()
functions <- list.files(here::here("functions"))

@@

tmdata <- used_data %>% 
  filter(targeted == 1) %>% 
  select(tm, everything())

median_tm = round(median(tmdata$tm))

@@

# Main Text

@@

Given this existing body of theory, what should we expect *a priori* the effect size of the Channel Islands MPAs to be? While we know many parameters will affect the expected effects of the MPAs, theory suggests that life history, adult and larval dispersal relative to MPA size, and fishing pressure will be key drivers. The targeted species in our database span a range of life histories, but are largely made up of fishes in the perch and rockfish complexes, with a mean Von Bertalanffy growth coefficient of `r round(mean(used_data$k[used_data$targeted == 1]),2)`, and a median age at maturity of `r round(median(used_data$tm[used_data$targeted == 1]),0)` years [values drawn from @thorson2017c].

@@

To illustrate this problem, we simulated response ratio trajectories for species and MPA coverage representative of the Channel Islands (see methods and Supporting Information (SI) Appendix for details of model structure). These simulations cover the life-histories and MPA sizes seen in the Channel Islands, but vary in key unknowns such as the actual degree of fishing pressure, the timing of density dependence, and the fishing fleets response to the MPAs. For each year of protection, we paired our simulated response ratios to the estimated posterior probability distributions of the  empirical Channel Island response ratios. For example, if the mean estimated response ratio in the year 2006 (year three after implementation) is one, we found all simulations that had simulated response ratios near one after three years of protection, and then pulled out the "true" simulated population-level MPA effect from each of those simulations. This provides us with a distribution of simulated population-level MPA effects that could plausibly generate the types of response ratios actually observed in the Channel Islands.

@@

@@

@@

While we do not have access to fine scale fishing data from the Channel Islands alone, reported catches for the species of interest in the Santa Barbara region exhibit a mix of stable, downward, and upward trajectories (see SI Appendix Fig.S1) which indicates that a negative MPA effect caused by a constant-catch fishing strategy is unlikely. However, some of the most common targeted species in the Channel Islands, such as blue rockfish and copper rockfish, have seen dramatic increases in reported catches in the Santa Barbara region since the year 2010 (although other such as California sheephead have seen declines relative to pre-MPA catch level). To the extent that these increases in catches are due to external market forces, the dissapearance of clear positive population-level MPA effects estimated by our model could be due to exogenous increases in fishing pressure overwhelming any population-level increases caused by the MPA network. We lack a control though that would allow us to estimate the MPA effects relative to a counter-factual in which the same market forces increase catches but no MPAs are in place.

@@

This hypothesis is supported by the fact that we estimate a similar decline in the estimated MPA effect both inside and outside of the MPAs (SI Appendix Fig.S18) (as well as observe the decline in the raw data, Fig.\@ref(fig:raw-density)). If the cause of recent declines was due to increases in fishing pressure we would expect to see substantial declines only in the fished areas. While we see evidence for the validity of the parallel trends assumption in the early years of our data (Fig.\@ref(fig:mpa-effect), SI Appendix Fig.S26), it is likely that the non-targeted species reacted in a different way to the marine heatwave than the targeted species did, causing a violation of the parallel trends assumption in the post 2014 data.

@@

Our simulations and difference-in-difference model examine percentage changes in biomass densities (as opposed to changes in total population size). An increase in biomass densities from 0.02kg/m^2^ to 0.04kg/m^2^ would be measured as a 100% increase. While a large percentage effect, this is a small change in biomass densities relative to the variance in the observation process itself. While an MPA providing 25% protection might be expected to produce a 30% increase in biomass densities, expressed as a percentage of unfished biomass, this gain is on the order of 15% (SI Appendix Fig.S45). In addition, the median estimated age at sexual maturity for the targeted species included in this study is `r median_tm` years, meaning that the span of this analysis represents less than three generations of MPA protection for half of the measured species. Ongoing monitoring may yet reveal clearer effects, especially if the time series extends beyond the periods of environmental shocks. Analysis of more rapidly growing and maturing species, e.g. spiny lobster, may also reveal clearer signals.

@@

When should we expect to see MPA effects big enough to stand a reasonable chance of detection? We simulated `r n_distinct(outcomes$experiment)` MPA scenarios across a wide range of life histories, network designs, and fishing dynamics. Suppose that we are willing to tolerate a MAPE of 25%. Our analysis suggests that we would need a population-level MPA effect size of at least 30% to achieve this. Two of the most critical drivers of MPA effect size are the size of the MPA network and the degree of fishing pressure. Looking across these two variables, if the MPA network covers 25% or more of a species range and pre-MPA depletion is greater than 60% we might expect an effect size with our target MAPE. Although recently some extremely large MPAs have been enacted that may indeed reach into the higher levels of MPA coverage, most MPA networks for near-shore commercial finfish are likely to cover areas more in line with the Channel Islands (20%) or smaller, and as such have population-level effect sizes that may be difficult to detect unless they were experiencing extreme overfishing prior to MPA establishment (Fig.\@ref(fig:sim-effects)-A).

@@

To put these results into context, the FAO estimates that 7% of the worlds fisheries with status estimates fall into the relatively unexploited category (roughly depletion less than 50%), 60% fall into the fully fished category (roughly depletion 50%-70%), and 33% fall into the heavily fished category (roughly depletion greater than 70%) [@fao2018, assuming a biomass at maximum sustainable yield to unfished biomass ratio of 40%, per @thorson2012], although studies that include a broader range of fisheries estimate that 50% or more of stocks fall into the heavily fished category [@costello2012;@rosenberg2018]. The population-level effects of MPA networks covering 25% or less of a species range may be difficult to detect in many places with already well-managed fisheries, while for that MPA network size we might expect clearer effects in less-managed locations (assuming compliance with MPA regulations). Within these broad guidelines a wide range of outcomes are possible based on local fleet and fish dynamics: users can use the bio-economic MPA simulation model developed for this paper to explore potential outcomes for specific MPAs using an interactive web application available [danovando.shinyapps.io/simmpa/](https://danovando.shinyapps.io/simmpa/).

@@

Non-equilibrium analyses also help set expectations for effect sizes over time [@nickols2019,@kaplan2019,Fig.\ref{did-plot}]. Educating communities about the challenges of estimating the effects of MPAs can help ensure that a lack of a clear effect is not necessarily viewed as a failure of the program, or large positive result based solely on response ratios as a clear sign of success and subsequent relaxation of other fishery management strategies. Rather, results and subsequent management actions (such as adaptation of MPA networks) must be considered in the context of reasonable expectations given the size, age, and degree of enforcement of the MPAs in question, together with the ecological and economic dynamics of a given system. While this paper has focused on the conservation outcomes of MPAs, future work must also address the challenge of predicting and estimating the fishery impacts of protected areas [in the manner of @smith2006a].

@@

We present here critical characteristics of our simulation model and regression approach. Further details can be found in the Supplementary Information. All analysis were conducted in R [@rcoreteam2019]. Our main difference-in-difference model was fit using Stan [@carpenter2017] using the `rstanarm` package [@goodrich2020]. All data and code needed to fully reproduce our results are publicly available at [github.com/DanOvando/population-effects-of-mpas](https://github.com/DanOvando/population-effects-of-mpas).

@@

Our bio-economic model simulates the effect of MPAs on a spatially explicit age-structured representation of a single species. Readers can explore the functionality of the model using an online tool available at [danovando.shinyapps.io/simmpa/](https://danovando.shinyapps.io/simmpa/). Full details of the simulation model are available in the Supplementary Information.

@@

@@

The key assumptions of the difference-in-difference model are that a) within the time-frame of the model there are no significant interaction effects between the targeted and non-targeted species (which in fact we do not detect, see SI Appendix), and b) that in the absence of the MPAs both the targeted and non-targeted groups of species would have exhibited similar trends in densities. It is important to note that the model is agnostic as to the root causes of the parallel trends, so long as they hold.

@@

where $d_i$ is the biomass density at observation *i*, *T* indicates whether the observation *i* is for a targeted ($T = 1$) or non-targeted ($T = 0$) species, and *MPA* marks whether observation *i* is in a pre MPA ($MPA = 0$) or post MPA ($MPA = 1$) state. **B^c^** is a vector of coefficients for additional control variables in matrix *X* such as water visibility and observer experience. **B^s^** is a vector of hierarchical coefficients for each sampling location **S** , clustered by island.  Under the assumptions of this model, $\beta_3$ is the causal effect of the treatment (*MPA*) on the treated (targeted species). $\beta_0$ is the mean total biomass of non-targeted species pre-MPA, and $\beta_1T$ is the mean total biomass of targeted species pre-MPA. $\beta_2MPA$ is the effect of the post-MPA period on the non-targeted species. The shape and scale parameters of the Gamma distribution are estimated as well. `r percent(mean(used_data$eventual_mpa))` of the samples in the raw data come from within MPAs, while the MPAs themselves cover 20% of the surface waters of the Channel Islands. As such, we weight each observation such that within MPA data are assigned a total weight of 0.2 and outside-MPA data a total weight of 0.8. See SI Appendix for sensitivity analyses of various model structures and synthetic controls results.

We briefly assess two of the most critical assumptions of this model here: that the treated and non-treated groups have parallel trends, and that the effect of the treatment on the treated does not tangentially affect the untreated. While the parallel trends assumption cannot be formally proven, we can examine its validity using the data from the years before the MPAs were put in place in 2003. We do not detect any significant differences in the trends of the biomass densities of the targeted and non-targeted species in the years before the MPAs (SI Appendix Fig.S26). However, we believe that the parallel trends assumption is likely violated in the post 2014 data.

With regards to the second assumption, all of the species in this empirical analysis exist within an ecosystem, and as such affect each other through mechanisms such as predation, competition, and habitat modification. We find it unlikely that these effects have had enough time to manifest in a meaningful way in the 14 years of post-MPA data used in our analysis [@babcock2010; @pershing2015a]. We used convergent cross mapping (CCM), in the manner of @clark2015, to test for the possibility of the trophic cascades biasing our results. Our results found no significant cross-mappings between targeted and non-targeted species, indicating that while clearly there are interactions between these groups on some level, the effects within the timespan of the data are not pronounced enough to be of concern to our results (SI Appendix Fig.S42). However, the longer MPAs are in place, the greater the possibility that substantial species interactions that can affect use of non-targeted species as a control may arise.

# Figures

ggsave(
  temp_plot,
  filename = here::here("documents","figs","response_ratio_plot.pdf"),
  device = cairo_pdf,
  width = width_1col,
  height = width_1col * asp_1col,
  units = "in",
)

ggsave(
  mpa_effect_plot,
  filename = here::here("documents","figs","mpa_effect_plot.pdf"),
  width = width_2col,
  height = width_2col * asp_2col,
  device = cairo_pdf
)

ggsave(
  temp_plot,
  filename = here::here("documents","figs","raw_biomass_density_plot.pdf"),
  width = width_1col,
  height = width_1col * asp_1col,
  device = cairo_pdf
)

temp_plot
ggsave(
  temp_plot,
  filename = here::here("documents","figs","validation_plot.pdf"),
  width = width_1col,
  height = width_1col * asp_1col,
  device = cairo_pdf
)

expected_mpa_effect_plot
ggsave(
  expected_mpa_effect_plot,
  filename = here::here("documents","figs","expected_mpa_effect.pdf"),
  width = width_2col,
  height = width_2col * asp_2col,
  device = cairo_pdf

temp_plot
ggsave(
  temp_plot,
  filename = here::here("documents","figs","sample_location_plot.pdf"),
  width = width_1col,
  height = width_1col * asp_1col,
  device = cairo_pdf
)

# References
